---
title: "Importing Raw Ascochyta Blight Lesion Count Data"
teaching: 60
exercises: 15
questions:
- "How can I prepare my data for analysis and save it without overwriting the raw data?"
objectives:
- " Understand how to import real-world data"
- " Calculate mean values from raw data"
- " Data frame manipulation including dropping and converting columns"
- " Save the newly formatted data for use in further work using R"
keypoints:
- "Use `readr` to import CSV files in place of `read.csv()`"
- "Use `dplyr` to add new columns of data."
- "Use `dplyr` to summarise data."
- "Create a data set that is useful for visualisation and modelling."
source: Rmd
---

```{r chunk_options, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("09-")
```

To get started with the Ascochyta blight (AB) data, we need to import it and  calculate mean lesion values.

In the first step, the data is imported, the "m" from the `distance` column is dropped and the column is converted to numeric.
A new column, `SpEv` (Spread Event), is created for each location and replicated event at that location. Another new column, `degrees`, is added for each transect to indicate the direction in degrees based on J. Fanning's best information.
Lastly the mean lesions counted per plant within the pot, `m_lesions` is calculated for each of the transects at each distance, 0, 10, 25, 50 and 75 meters from the plots.

After the data are updated, a column for precipitation type, `ptype`, is added and the spread events, `SpEv`, are renamed for clarity and ordered as factors for data visualisation.

## Load libraries

We will use the `tidyverse` to import and format these data for use in our model.

```{r libraries}
library("tidyverse")
```

## Import and Inspect Lesion Data

Using `read_csv()` import the lesion count data from Zenodo where they are archived.
See [10.5281/zenodo.3842293](https://doi.org/10.5281/zenodo.3842293) for a full description of the data.

```{r import_data, message=FALSE}
lesion_counts <-
   read_csv("https://zenodo.org/record/3842293/files/lesion_counts.csv?download=1")
```


```{r}
   mutate(site = str_remove(site, " SPA")) %>%
   mutate(distance = as.numeric(str_replace(distance, " m", ""))) %>%
   unite(SpEv, c(site, rep), remove = FALSE) %>%
   mutate(
      degrees = case_when(
         site == "Curyo" & transect == 1 ~ 290,
         site == "Curyo" & transect == 2 ~ 300,
         site == "Curyo" & transect == 3 ~ 310,
         site == "Curyo" & transect == 4 ~ 320,
         site == "Curyo" & transect == 5 ~ 330,
         site == "Curyo" & transect == 6 ~ 340,
         site == "Curyo" & transect == 7 ~ 350,
         site == "Curyo" & transect == 8 ~ 360,
         site == "Curyo" & transect == 9 ~ 10,
         site == "Curyo" & transect == 10 ~ 20,
         site == "Horsham" & transect == 1 ~ 45,
         site == "Horsham" & transect == 2 ~ 55,
         site == "Horsham" & transect == 3 ~ 65,
         site == "Horsham" & transect == 4 ~ 75,
         site == "Horsham" & transect == 5 ~ 85,
         site == "Horsham" & transect == 6 ~ 95,
         site == "Horsham" & transect == 7 ~ 105,
         site == "Horsham" & transect == 8 ~ 115,
         site == "Horsham" & transect == 9 ~ 125,
         site == "Horsham" & transect == 10 ~ 135,
         site == "pbc" & transect == 1 ~ 45,
         site == "pbc" & transect == 2 ~ 55,
         site == "pbc" & transect == 3 ~ 65,
         site == "pbc" & transect == 4 ~ 75,
         site == "pbc" & transect == 5 ~ 85,
         site == "pbc" & transect == 6 ~ 95,
         site == "pbc" & transect == 7 ~ 105,
         site == "pbc" & transect == 8 ~ 115,
         site == "pbc" & transect == 9 ~ 125,
         site == "pbc" & transect == 10 ~ 135
      )
   ) %>%
   mutate(
      ptype = case_when(
         SpEv == "pbc_1" ~ "irrigation",
         SpEv == "pbc_2" ~ "irrigation",
         SpEv == "pbc_3" ~ "mixed",
         SpEv == "Horsham_1" ~ "rainfall",
         SpEv == "Horsham_2" ~ "rainfall",
         SpEv == "Curyo_1" ~ "rainfall"
      )
   ) %>%
   mutate(
      site = case_when(
         site == "pbc" ~ "Horsham irrigated",
         site == "Horsham" ~ "Horsham dryland",
         site == "Curyo" ~ "Curyo"
      )
   ) %>%
   mutate(
      SpEv = case_when(
         SpEv == "pbc_1" ~ "Horsham Irrg 1",
         SpEv == "pbc_2" ~ "Horsham Irrg 2",
         SpEv == "pbc_3" ~ "Horsham Mixd 1",
         SpEv == "Horsham_1" ~ "Horsham Rain 1",
         SpEv == "Horsham_2" ~ "Horsham Rain 2",
         SpEv == "Curyo_1" ~ "Curyo Rain 1"
      )
   ) %>%
   mutate_at(vars(SpEv, site, rep), factor) %>%
   mutate(m_lesions = rowMeans(select(., counts_p1:counts_p5),
                                    na.rm = TRUE)) %>% 
   select(-dist_stat) %>% 
   arrange(SpEv, transect, distance)

kable(lesion_counts)
```
