---
title: "Fit and Compare Generalised Additive Models of Conidial Dispersal"
teaching: 45
exercises: 15
questions:
- "How can we use a generalised additive model (GAM) to represent conidial dispersal?"
- "What can the GAM tell us about the data?"
objectives:
- "Join different sets of data together for building a model."
- "Fit generalised additive models to data."
- "Determine which factors are necessary to include in the model."
- "Evaluate different models to determine which model provides the best fit."
- "Visualise a generalised additive model."
keypoints:
- "Generalised additive models are a powerful way to model non-linear relationships."
- "When fitting models to data, check many ways and compare."
- "There are no perfect models, only some that are better than others."
source: Rmd
---

```{r chunk_options, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("11-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9.4, fig.height = 9.4)
```

In this episode we will explore how to fit and evaluate the best fitting model for our data.
I've chosen to use a generalised additive model for this work because as we saw in the data visualisation episode, the lesion count data is not linear, *i.e.* the lesion counts follow a curve not a straight line as you get farther from the source of inoculum.

### Load Libraries for This Exercise

Loading the `ChickpeaAscoDispersal` library will load the data we previously generated.

```{r libraries, message=FALSE, echo=FALSE}
library("ChickpeaAscoDispersal")
library("tidyverse")
library("broom") # provides a way to "tidy" model outputs
library("mgcv") # provides GAM fits
library("mgcViz") # use ggplot2 to visulalise GAMs

theme_set(theme_bw(base_size = 14))
```

## Create Data Set for GAMs

Use `dplyr::left_join()` join the `lesion_counts` data and the `summary_weather` data to create `dat` for creating GAMs.
`left_join()` is a powerful tool for combining data sets based on one or more matching columns.

```{r import_data, message=FALSE}
dat <-
   left_join(lesion_counts, summary_weather, by = c("site", "rep"))
```

```{r fonts, include=FALSE, message=FALSE, eval=FALSE}
# run only if knitting on new computer/new R installation
# !!!! Run this in a regular interactive console not this markdown file !!!!
# this is necessary to embed fonts in .eps files for EJPP
# run only if knitting on new computer
# this is necessary to embed fonts in .eps files for EJPP
library("extrafont")

if (.Platform$OS.type == "windows") {
   font_import(pattern = "arial", prompt = FALSE)
   loadfonts(device = "postscript", quiet = TRUE)
} else {
   font_import(pattern = "Arial", prompt = FALSE)
   loadfonts(device = "postscript", quiet = TRUE)
}
```

## Fit GAMs

For reproducibility purposes, use `set.seed()`.

```{r set-seed, echo=TRUE}
set.seed(27)
```

### mod1 - s(Distance)

```{r fit-mod1}
mod1 <-
   gam(m_lesions ~ s(distance, k = 5),
       data = dat)

summary(mod1)

print(p_gam(x = getViz(mod1)) +
         ggtitle("s(Distance)"),
      pages = 1)

mod1_vis <- getViz(mod1)
check(
   mod1_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod2 - s(Distance) + Precipitation

```{r fit-mod2}
mod2 <-
   gam(m_lesions ~ sum_rain + s(distance, k = 5),
       data = dat)

summary(mod2)

print(p_gam(x = getViz(mod2)) +
         ggtitle("s(Distance) + Precipitation"),
      pages = 1)

mod2_vis <- getViz(mod2)
check(
   mod2_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod3 - s(Distance) + Wind speed

```{r fit-mod3}
mod3 <-
   gam(m_lesions ~ mws + s(distance, k = 5),
       data = dat)

summary(mod3)

print(p_gam(x = getViz(mod3)) +
         ggtitle("s(Distance) + Wind speed"),
      pages = 1)

mod3_vis <- getViz(mod3)
check(
   mod3_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod4 - s(Distance) + Wind speed + Precipitation
```{r fit-mod4}
mod4 <-
   gam(m_lesions ~ sum_rain + mws + s(distance, k = 5),
       data = dat)

summary(mod4)

print(p_gam(x = getViz(mod4)) +
         ggtitle("s(Distance) + Wind speed + Precipitation"),
      pages = 1)

mod4_vis <- getViz(mod4)
check(
   mod4_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod5 - s(Distance + Wind Speed) + Precipitation

```{r fit-mod5}
mod5 <-
   gam(m_lesions ~ sum_rain + s(distance + mws, k = 5),
       data = dat)

summary(mod5)


print(p_gam(x = getViz(mod5)) +
         ggtitle("s(Distance + Wind Speed) + Precipitation"),
      pages = 1)

mod5_vis <- getViz(mod5)
check(
   mod5_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod6 - s(Distance) + s(Wind Speed) + Precipitation

```{r fit-mod6}
mod6 <-
   gam(m_lesions ~ sum_rain + s(distance, k = 5) + s(mws, k = 5),
       data = dat)

summary(mod6)


print(p_gam(x = getViz(mod6)) +
         ggtitle("s(Distance) + s(Wind Speed) + Precipitation"),
      pages = 1)

mod6_vis <- getViz(mod6)
check(
   mod6_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod7 - s(Distance) + s(Wind Speed)

```{r fit-mod7}
mod7 <-
   gam(m_lesions ~ s(distance, k = 5) + s(mws, k = 5),
       data = dat)

summary(mod7)

print(p_gam(x = getViz(mod7)) +
         ggtitle("s(Distance) + s(Wind Speed)"),
      pages = 1)

mod7_vis <- getViz(mod7)
check(
   mod7_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod8 - s(Distance) + s(Wind Speed) + s(Precipitation)

```{r fit-mod8}
mod8 <-
   gam(m_lesions ~ s(distance, k = 5) + s(mws, k = 5) + s(sum_rain, k = 5),
       data = dat)

summary(mod8)

print(p_gam(x = getViz(mod8)) +
         ggtitle("s(Distance) + s(Wind Speed) + s(Precipitation)"),
      pages = 1)

mod8_vis <- getViz(mod8)
check(
   mod8_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod9 - s(Distance) + s(Precipitation)

```{r fit-mod9}
mod9 <-
   gam(m_lesions ~ s(distance, k = 5) + s(sum_rain, k = 5),
       data = dat)

summary(mod9)

print(p_gam(x = getViz(mod9)) +
         ggtitle("s(Distance) + s(Precipitation)"),
      pages = 1)

mod9_vis <- getViz(mod9)
check(
   mod9_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod10 - s(Distance) +s(Precipitation) + Wind speed

```{r fit-mod10}
mod10 <-
   gam(m_lesions ~ s(distance, k = 5) + s(sum_rain, k = 5) + mws,
       data = dat)

summary(mod10)

print(p_gam(x = getViz(mod10)) +
         ggtitle("s(Distance) + s(Precipitation) + Wind speed"),
      pages = 1)

mod10_vis <- getViz(mod10)
check(
   mod10_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod11 - s(Distance) + s(Wind Speed) + s(Precipitation), family = tw()

This is the same as `mod8` but using `family = tw()`, see `?family.mgcv` for more on the families.
The Tweedie distribution is used where the distribution has a positive mass at zero, but is continuous unlike the Poisson distribution that requires count data.
The data visualisation shows clearly that the mean pot count data have this shape.

```{r fit-mod11}
mod11 <-
   gam(
      m_lesions ~ s(distance, k = 5) +
         s(mws, k = 5) +
         s(sum_rain, k = 5),
      data = dat,
      family = tw()
   )

summary(mod11)

print(
   p_gam(x = getViz(mod11)) +
      ggtitle("s(Distance) + s(Wind Speed) + s(Precipitation), family = tw()"),
   pages = 1
)

mod11_vis <- getViz(mod11)
check(
   mod11_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod12 - s(Distance, bs = "ts") + s(Precipitation, bs = "ts") Wind speed, family = tw()

Try using wind speed as a linear predictor only.

```{r fit-mod12}
mod12 <-
   gam(
      m_lesions ~ s(distance, k = 5, bs = "ts") +
         s(mws, k = 5, bs = "ts") +
         s(sum_rain, k = 5, bs = "ts"),
      data = dat,
      family = tw()
   )

summary(mod12)

print(
   p_gam(x = getViz(mod12)) +
      ggtitle(
         "s(Distance, bs = 'ts') + s(Wind speed, bs = 'ts')\n+ s(Precipitation, bs = 'ts'), family = tw()"
      ),
   pages = 1
)

mod12_vis <- getViz(mod12)
check(
   mod12_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

### mod13 - s(Distance, bs = "ts") + s(Wind speed, bs = "ts") + s(Precipitation, bs = "ts"), family = tw()

```{r fit-mod13}
mod13 <-
   gam(
      m_lesions ~ s(distance, k = 5, bs = "ts") +
         s(mws, k = 5, bs = "ts") +
         s(sum_rain, k = 5, bs = "ts"),
      data = dat,
      family = tw()
   )

summary(mod13)

print(
   p_gam(x = getViz(mod13)) +
      ggtitle(
         "s(Distance, bs = 'ts') + s(Wind speed, bs = 'ts')\n+ s(Precipitation, bs = 'ts'), family = tw()"
      ),
   pages = 1
)

mod13_vis <- getViz(mod13)
check(
   mod13_vis,
   a.qq = list(method = "tnorm",
               a.cipoly = list(fill = "light blue")),
   a.respoi = list(size = 0.5),
   a.hist = list(bins = 10)
)
```

This model, same structure as `mod11`, uses thin-plate splines to shrink the coefficients of the smooth to zero when possible.

## Compare the Models

### AIC, BIC
```{r compare-mods}
models <- list(
   mod1 = mod1,
   mod2 = mod2,
   mod3 = mod3,
   mod4 = mod4,
   mod5 = mod5,
   mod6 = mod6,
   mod7 = mod7,
   mod8 = mod8,
   mod9 = mod9,
   mod10 = mod10,
   mod11 = mod11,
   mod12 = mod12,
   mod13 = mod13
)
map_df(models, glance, .id = "model") %>%
   arrange(AIC)
```

### R^2^
```{r r2}
enframe(
   c(
      mod1 = summary(mod1)$r.sq,
      mod2 = summary(mod2)$r.sq,
      mod3 = summary(mod3)$r.sq,
      mod4 = summary(mod4)$r.sq,
      mod5 = summary(mod5)$r.sq,
      mod6 = summary(mod6)$r.sq,
      mod7 = summary(mod7)$r.sq,
      mod8 = summary(mod8)$r.sq,
      mod9 = summary(mod9)$r.sq,
      mod10 = summary(mod10)$r.sq,
      mod11 = summary(mod11)$r.sq,
      mod12 = summary(mod12)$r.sq,
      mod13 = summary(mod13)$r.sq
   )
) %>%
   arrange(desc(value))
```

### ANOVA
```{r anova}
anova(mod1,
      mod2,
      mod3,
      mod4,
      mod5,
      mod6,
      mod7,
      mod8,
      mod9,
      mod10,
      mod11,
      mod12,
      mod13,
      test = "F")
```

> ## Challenge 1
>
> Discuss with your neighbours which model is the best fit and why.
>
> What are the criteria that you've used to determine this?
>
> What can you say about the model?
> 
> What can you NOT say about the model?
>
> > ## Solution to Challenge 1
> >This model, _mod11_,
> > `m_lesions ~ s(Distance) + s(WindSpeed) + s(Precipitation) - family = tw()`,
> >is the best performing model.
> >
> >*It cannot be used for predictions*, but it suitably describes the dispersal
> >data we have on hand with the parameters used.
> >
> >More data would be desirable to increase the value of `k` as evidenced in the
> >GAM checks.
> >
> {: .solution}
{: .challenge}
