---
title: "Check if Location Affects Conidial Dispersal"
teaching: 60
exercises: 15
questions:
- "Did experimental location (site) affect conidial dispersal?"
- "?"
- "?"
objectives:
- ""
keypoints:
- ""
source: Rmd
---

```{r chunk_options, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("11-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 9.4, fig.height = 9.4)
```

In this experiment there are a number of factors which may influence conidia spread.
These include:

  * Experimental location (site).

  * The time at which a spread event occurred (`SpEv`).

    - The factor `SpEv` would be nested within site.

    - `SpEv` factor may describe variation in the data that varies between `SpEv`, such as weather and climate variables.
  
  * Wind speed during the spread event.

  * Wind direction during the spread event.

  * Distance the trap plants were placed from the aschochyta infested plots (distance).

  * The bearing in which the trap plants were placed at distances relative to the infested plots (transect).

  * How the spread event was initiated, with sprinkler irrigation or rainfall.

  * The quantity of rainfall.

Due to the lack of replicated pots at some of the distances we will ignore transect as a factor.
We know wind direction will influence our results and we will need to accept that adds variation for which we may not be able to account for statistically.

I will start using `lmer()` to analyse the mean number of lesions per plant at each distance.
The reps at each distance are defined by 'pot', each pot contains three to five chickpea plants.
The factor distance is fit as a continuous variable.  

Site is a categorical variable explaining the trial location.
Each site may have experienced a different number of spread events, defined by the term `SpEv`.
Rainfall is required for conidia to disperse from the infected focus, and each 'spread event' constitutes either an overhead irrigation event or a natural rainfall event.

The first models we will look at are asking:

   (i) What is the estimated mean lesions per plant as each distance from the focus, given that the conditions of each spread event is nested within each site and the distance the spore travels is dependent on the spread event at each site.
   (ii) What is the estimated mean lesions per plant as each distance from the focus, given that distance is dependant on the conditions of each spread event.

### Load Libraries for This Exercise

```{r libraries, message=FALSE, warning=FALSE}
library("ChickpeaAscoDispersal")
library("tidyverse")
library("lme4")
```

## Model Fitting

```{r lmer_mod1, message=FALSE}
dat <-
  left_join(lesion_counts, summary_weather, by = c("site", "rep"))

mod1 <-
  lmer(m_lesions ~ distance + (distance | site / SpEv),
       data = dat)

cat("mod1: ")
formula(mod1)

summary(mod1)
```

Let's examine the model without `site` to test if the model is a worse fit.

```{r lmer_mod2}
mod2 <-
  lmer(m_lesions ~ distance +
         (distance | SpEv),
       data = dat)

cat("mod2: ")
formula(mod2)

# Compare models
anova(mod1, mod2)
```

A comparison of the two models shows us that `mod2` is much better fit given the lower AIC and that there is no significant difference in the models. Following a reductive approach we should remove site from the model.

```{r mod2}
summary(mod2)
```

We can also note that as the distance increases there are less mean lesions per pot, and the variance increases.

From here we should continue with a generalised additive model (GAM), which can handle non-linear terms better than a linear model.
