---
title: "Working with Spatial Data in R"
teaching: 10
exercises: 10
questions:
- "What is special about spatial data?"
- "What packages are available for making maps in R?"
objectives:
- "Understand what the differences in spatial data format are"
- "Gain some familiarity with R packages that enable working with spatial data in R"
keypoints:
- "R has a broad ecosystem of spatial packages to use."
- "Understand differences in spatial data, raster is for continous data, vector is for discreet data."
source: Rmd
---

```{r chunk_options, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("18-")
```

## Forms of Spatial Data

Spatial data comes in two basic forms, raster (gridded) and vector (points, lines and polygons).

Raster data are often used to represent continuous variables, _e.g._ temperature, rainfall or elevation. Whereas vector data are used to represent sampling locations, weather station locations (and data at that point), roads, state or national outlines or other regions that may be natural or manufactured.

Understanding how these data differ is important to being able to use them in maps in R.

## Spatial Packages

Spatial data has its own unique properties. Packages that are commonly used with spatial data in R include.

* [_raster_](https://CRAN.R-project.org/package=raster),
* [_terra_](https://CRAN.R-project.org/package=terra),
* [_sf_](https://CRAN.R-project.org/package=raster),
* [_rnaturalearth_](https://CRAN.R-project.org/package=rnaturalearth),
* [_rnaturalearthdata_](https://CRAN.R-project.org/package=rnaturalearthdata),
* [_ggspatial_](https://CRAN.R-project.org/package=ggspatial),
* _grid_ (this is a part of your base R installation) and
* [_gridExtra_](https://CRAN.R-project.org/package=gridExtra).

The _raster_ package is used to handle raster data files but also offers capabilities to download and import other data including country-level outlines ([GADM](https://gadm.org/)), elevation (SRTM data) and bioclimatic variables from [Worldclim](https://www.worldclim.org/).

The _terra_ package is a replacement for the _raster_ package developed by the same person to be faster and more simple.

_sf_ is used to handle vector-format files and _ggplot2_ now offers `geom_sf()` for easy plotting of _sf_ objects.

_rnaturalearth_ and _rnaturalearthdata_ facilitate mapping using data from [http://www.naturalearthdata.com/](http://www.naturalearthdata.com/).

_ggspatial_ is used to add map elements such as scale bars and north arrows to _ggplot2_ objects.

_grid_ and _gridExtra_ are used to add what are called "neatlines", an outline around the map to contain all the elements of the map.

You may also want to visit [https://rspatial.org](https://rspatial.org/) for more packages, tutorials and information.


> ## Challenge 1
>
> Install the spatial packages necesary for working with spatial data in R.
> They are terra, sf, rnaturalearth, rnaturalearthdata and ggspatial.
>
> > ## Solution to Challenge 1
> >```{r}
> > install.packages(c(
> >   "raster",
> >   "sf",
> >   "rnaturalearth",
> >   "rnaturalearthdata",
> >   "ggspatial"
> > ))
> > ```
> {: .solution}
>
{: .challenge}


Once the main packages are installed, the second step is to install the _rnaturalearthhires_ package as well.

Answer `1` to download and install the necessary data. Again, this will also take some time.

```{r install-spatial-2, eval=FALSE}
library("rnaturalearth")
install_rnaturalearthhires()
```

Now we are ready to make some maps!
