---
title: "Creating Country and State Level Maps"
teaching: 15
exercises: 10
questions:
- "How can I get outlines of countries or states in R?"
- "How do I highlight a specific state?"
objectives:
- "Understand how to retreive data for making maps."
- "Understand how to manipulate and subset data to create maps."
keypoints:
- "R has packages that make it easy to retrieve data and create maps."
- "The sf package works nicely with ggplot2 and tidyverse."
source: Rmd
---


```{r chunk_options, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("19-")
```

In this first exercise we will use [_rnaturalearth_](https://github.com/ropenscilabs/rnaturalearth), [_simple features_](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html) and
_ggplot2_ to create a map of India and states within India to use as the base of the maps we will create in these exercises.

## Create Country Level Map of India

This is our base layer, India, of the map from [Naturalearth.com](https://naturalearth.com/) using the base `plot()` to display these data.

Using `ne_states()` will return an `sf` object with state level information. There is also a `ne_countries()` to download country level outlines only as well.

```{r India-base}
library("rnaturalearth")

in_sf <- ne_states(geounit = "india",
                   returnclass = "sf")

plot(in_sf)
```

### Plot using `ggplot2`

Using _ggplot2_, we have more control over how the data are displayed.

First plot the whole country using `ggplot()`.

```{r plot_level1, eval=TRUE}
library("ggplot2")

india <- ggplot(in_sf) +
  geom_sf(fill = "white")

india
```

### Add Labels to the States

The naturalearth data have several columns to work with. One of them is the abbreviation for the states. Using `geom_text()`, we can add this information to the map. 

```{r plot_level2}
india <-
  india +
  geom_text(
    data = in_sf,
    aes(x = longitude,
        y = latitude,
        label = postal),
    size = 2.5,
    hjust = 1
  )

india
```

### Final Touches

Properly label the x and y-axis and set the theme. The _ggplot2_ theme, `theme_bw()` is nice to use with maps so we will apply that here as well.

```{r plot_level3}
india <- 
  india +
  xlab("Longitude") +
  ylab("Latitude") +
  theme_bw()

india
```

## Create Maps of Individual States

To create a map of only Andhra Pradesh, subset the _sf_ object, `india` such that it contains only the data for Andhra Pradesh.

### Inspect Column Names in `in_sf` Object

To find out what the column names are, we can use, `colnames()`.

```{r colnames}
names(in_sf)
```

We see that a column named "name" exists, with the name of each state.

We can use this to `filter()` the data so that it only represents Andhra Pradesh and we can use that to make a map.

For more on subsetting _sf_ objects, see: <https://cran.rstudio.com/web/packages/sf/vignettes/sf4.html>

### Filter Andhra Pradesh

```{r subset-ap}
library("dplyr")

ap <- filter(in_sf, name == "Andhra Pradesh")

ap
```

### Plot Andhra Pradesh

```{r ggplot-ap}
ggplot(data = ap) +
  geom_sf() +
  xlab("Longitude") +
  ylab("Latitude") +
  theme_bw()
```

## Highlighting States Within the Country

Using the `in_sf` and `ap` objects that we have created, it is possible to create a map that highlights one state and labels it using the abbreviation, in this case Andhra Pradesh.

```{r hightlight-ap}
ggplot(data = in_sf) +
  geom_sf(fill = "white") +
  geom_sf(
    data = ap,
    fill = "red") +
  geom_text(
    data = ap,
    aes(x = longitude,
        y = latitude,
        label = postal),
    size = 2.5,
    hjust = 1) +
  xlab("Longitude") +
  ylab("Latitude") +
  theme_bw()
```


> ## Challenge 1
>
> Create a map of India that highlights another state of your choice and label
> only that state.
> Make the state a different colour other than red.
>
> > ## Solution to Challenge 1
> >```{r}
> > tn <- filter(in_sf, name == "Tamil Nadu")
> > 
> > ggplot(data = in_sf) +
> >  geom_sf(fill = "white") +
> >  geom_sf(
> >    data = tn,
> >    fill = "yellow") +
> >  geom_text(
> >    data = tn,
> >    aes(x = longitude,
> >        y = latitude,
> >        label = postal),
> >    size = 2.5,
> >    hjust = 1) +
> >  xlab("Longitude") +
> >  ylab("Latitude") +
> >  theme_bw()
> > ```
> {: .solution}
>
{: .challenge}

